---
title: "edstr_clean(): text cleaning"

vignette: >
  %\VignetteIndexEntry{edstr_clean(): text cleaning}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
library(edstr)
```

`edstr_clean()` applies regex-based replacements to a text column and saves the cleaned data as an RDS file.
It requires `edstr_config()` to be called first.

## Prerequisites

```{r}
edstr_config(
  edstr_dirname = "output/my_study",
  edstr_filename = "my_study",
  edstr_text = "note_text",
  edstr_overwrite = FALSE
)
```

## Defining replacement rules

The `replace` argument takes a named character vector where names are regex patterns and values are their replacements.
Replacements are applied sequentially via `stringr::str_replace_all()`.

```{r}
#| error: true
df_clean <- edstr_clean(
  data = df_import,
  replace = c(
    "\\p{Zs}{2,}" = " ",
    "&amp;"        = "&",
    "&lt;"         = "<",
    "&gt;"         = ">"
  )
)
```

## Using a list for ordered replacements

When the order matters (e.g. fixing HTML entities before stripping tags), pass a list of named vectors.
Each element is applied in sequence.

```{r}
#| error: true
df_clean <- edstr_clean(
  data = df_import,
  replace = list(
    entities = c("&amp;" = "&", "&lt;" = "<", "&gt;" = ">"),
    whitespace = c("\\p{Zs}{2,}" = " ", "\\n+" = " ")
  )
)
```

## Externalising cleaning rules

For complex or project-specific patterns, define replacements in a standalone R file and load them with `source()`.
This keeps the cleaning logic versioned, testable, and reusable across scripts.

```{r}
#| error: true
df_clean <- edstr_clean(
  data = df_import,
  replace = source(here::here("config/clean.R"))$value
)
```

The sourced file should return a named character vector or list — see `demo/config/clean.R` for an example.

## Specifying the text column

By default, `edstr_clean()` uses the `edstr_text` option set in `edstr_config()`.
To override it for a specific call:

```{r}
#| error: true
df_clean <- edstr_clean(
  data = df_import,
  text = "other_column",
  replace = c("\\s+" = " ")
)
```

## Caching

Like other pipeline functions, `edstr_clean()` saves its output as `{filename}_clean.rds`.
On subsequent calls, it loads from cache based on the `edstr_overwrite` setting.
