---
title: "edstr_config(): pipeline configuration"

vignette: >
  %\VignetteIndexEntry{edstr_config(): pipeline configuration}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
library(edstr)
```

`edstr_config()` sets global options that all downstream functions rely on.
It must be called before any other `edstr_*` function.

## Required arguments

Two arguments are mandatory: the output directory and the file name prefix.

```{r}
edstr_config(
  edstr_dirname = "output/my_study",
  edstr_filename = "my_study"
)
```

The directory is created automatically if it does not exist.
Output files are named `{edstr_filename}_{step}.rds` — e.g. `my_study_import.rds`, `my_study_clean.rds`.

## Glue syntax

Both `edstr_dirname` and `edstr_filename` support [glue](https://glue.tidyverse.org/) expressions, which are evaluated at call time.

```{r}
edstr_config(
  edstr_dirname = "output/{Sys.Date()}",
  edstr_filename = "study_{format(Sys.Date(), '%Y%m%d')}"
)
```

## Text column

Setting `edstr_text` avoids repeating the column name in `edstr_clean()`, `edstr_extract()`, and `edstr_view()`.

```{r}
edstr_config(
  edstr_dirname = "output/my_study",
  edstr_filename = "my_study",
  edstr_text = "note_text"
)
```

If omitted, the text column must be passed explicitly to each function.

## Caching behaviour

The `edstr_overwrite` option controls what happens when an output file already exists.

| Value   | Behaviour |
|---------|-----------|
| `TRUE`  | Overwrite without prompting |
| `FALSE` | Load the cached file without prompting |
| `NULL`  | Prompt an interactive menu (load / overwrite / cancel) |

`NULL` is the default — useful during interactive development.
Set `TRUE` when re-running a full pipeline, `FALSE` when resuming work on a cached dataset.

```{r}
edstr_config(
  edstr_dirname = "output/my_study",
  edstr_filename = "my_study",
  edstr_overwrite = TRUE
)
```

## Additional options

Extra named arguments are passed directly to `options()`, which can be useful for setting project-level options alongside `edstr`.

```{r}
edstr_config(
  edstr_dirname = "output/my_study",
  edstr_filename = "my_study",
  dplyr.summarise.inform = FALSE
)
```
