---
title: "Get started"

vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
library(edstr)
```

The `{edstr}` package extracts structured variables from clinical free text stored in a Clinical Data Warehouse (EDS).
The pipeline runs in four steps: **configure**, **import**, **clean**, **extract**.
A fifth function, `edstr_view()`, lets you explore matches interactively without saving any files.

## Pipeline overview

```{mermaid}
%%| label: fig-pipeline
%%| fig-cap: "edstr pipeline"
flowchart LR
  A["edstr_config()"] --> B["edstr_import()"]
  B --> C["edstr_clean()"]
  C --> D["edstr_extract()"]
  C -.-> E["edstr_view()"]
```

Each step (except `edstr_view()`) saves an `.rds` file in the output directory.
If the file already exists, behaviour depends on the `edstr_overwrite` option set in `edstr_config()`.

## Configuration

`edstr_config()` sets shared parameters used by all downstream functions.
Two arguments are required: the output directory and the file name prefix.

```{r}
#| eval: false
edstr_config(
  edstr_dirname = "output/fesf",
  edstr_filename = "fesf",
  edstr_text = "doc_texte",
  edstr_overwrite = TRUE
)
```

`edstr_dirname` and `edstr_filename` support [glue](https://glue.tidyverse.org/) syntax, e.g. `"output/{Sys.Date()}"`.

::: {.callout-tip}
## Caching behaviour

- `edstr_overwrite = TRUE`: overwrite without prompting.
- `edstr_overwrite = FALSE`: load the cached file without prompting.
- `edstr_overwrite = NULL` (default): prompt an interactive menu (load / overwrite / cancel).
:::

## Import

`edstr_import()` runs a SQL query against the Oracle database and saves the result.

```{r}
#| eval: false
df_import <- edstr_import(
  query = "sql/fesf.sql",
  head = 500,
  user = "my_user"
)
```

The `query` argument accepts either a SQL string or a path to a `.sql` file.
`head` limits the number of returned rows — useful during development.

## Cleaning

`edstr_clean()` applies regex replacements to the text column.
Cleaning patterns are defined as a named character vector (or a list of named vectors): names are patterns, values are replacements.

```{r}
#| eval: false
df_clean <- edstr_clean(
  data = df_import,
  replace = c("\\p{Zs}{2,}" = " ", "&amp;" = "&")
)
```

::: {.callout-tip}
## Reusable cleaning patterns

For complex patterns, define them in a separate R file and load them with `source()`:

```{r}
#| eval: false
df_clean <- edstr_clean(
  data = df_import,
  replace = source(here::here("config/clean.R"))$value
)
```
:::

## Extraction

`edstr_extract()` is the core of the pipeline.
From cleaned text, it tokenises, matches concepts (regex), applies false-positive exclusions, and saves results as XLSX, CSV, and RDS files.

### Concepts

Concepts are named regex patterns defining the clinical entities to search for.
A simple vector creates independent concepts; a named list creates sub-concepts.

```{r}
#| label: extract-simple
#| eval: false
df_extract <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture", femur = "f(e|é)mur|fesf"),
  token = c(1, 2),
  group = "id_pat"
)
```

### Collapse and intersect

Two modes alter extraction behaviour when multiple concepts are defined:

- `collapse = TRUE`: combines all patterns into a single regex (logical OR).
- `intersect = TRUE`: keeps only documents matching **all** root concepts.

```{r}
#| label: extract-intersect
#| eval: false
df_extract <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture", femur = "f(e|é)mur|fesf"),
  intersect = TRUE,
  group = "id_pat"
)
```

### Exclusions

False positives can be filtered in two ways:

- **Manual**: a regex pattern via `exclus_manual`.
- **Automatic**: heuristics on long tokens (controlled by `exclus_auto_token_min`).

```{r}
#| label: extract-exclusions
#| eval: false
df_extract <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  exclus_manual = "fracture ouverte",
  exclus_auto_token_min = 10,
  group = "id_pat"
)
```

## Interactive exploration

`edstr_view()` lets you test regex patterns without running a full extraction.
This function does not save anything — it is designed for fast iteration.

```{r}
#| eval: false
df_view <- edstr_view(
  data = df_clean,
  pattern = "fractur",
  ngrams = 3
)
```

`ngrams` controls the number of tokens captured after the initial match: `ngrams = 3` with `pattern = "fractur"` matches e.g. `"fracture col femoral"`.

## Output summary

`edstr_extract()` returns a nested list and saves three files:

| File    | Contents |
|---------|----------|
| `.xlsx` | Excel workbook with one sheet per result type (extraction, counts, exclusions, mismatch, parameters) |
| `.csv`  | Flat extraction table, ready for analysis |
| `.rds`  | Full list with all intermediate objects |

The `.rds` file is the one automatically reloaded by the caching system when it already exists.
