---
title: "edstr_view(): interactive exploration"

vignette: >
  %\VignetteIndexEntry{edstr_view(): interactive exploration}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
library(edstr)
```

`edstr_view()` searches for regex patterns in a text column and displays matching tokens, counts, and highlighted text.
Unlike other pipeline functions, it does not save any files — it is designed for fast, iterative pattern development before running `edstr_extract()`.

## Prerequisites

```{r}
#| error: true
edstr_config(
  edstr_dirname = "output/my_study",
  edstr_filename = "my_study",
  edstr_text = "note_text",
  edstr_overwrite = FALSE
)

df_clean <- edstr_clean()
```

## Basic usage

Pass a regex pattern to search for in the text column.

```{r}
#| error: true
result <- edstr_view(
  data = df_clean,
  pattern = "fractur"
)
```

The function prints a frequency table of distinct matches and a summary of match counts, then invisibly returns a list with three elements:

- `match`: a tibble of all matches (one row per match, with the document ID)
- `count`: a tibble of distinct matches sorted by frequency
- `text`: highlighted output from `stringr::str_view()`

## Capturing multi-word expressions

The `ngrams` argument controls how many tokens are captured after the initial match.
This is useful for discovering common multi-word patterns.

```{r}
#| error: true
result <- edstr_view(
  data = df_clean,
  pattern = "fractur",
  ngrams = 3
)
```

With `ngrams = 3`, a match on `"fractur"` captures up to two additional tokens — e.g. `"fracture col femoral"` or `"fracture extremite superieure"`.

## Applying cleaning rules inline

The `replace` argument accepts the same format as `edstr_clean()` and applies replacements before matching.
This lets you test cleaning rules without modifying the data.

```{r}
#| error: true
result <- edstr_view(
  data = df_clean,
  pattern = "diabete",
  replace = c("diabète" = "diabete"),
  ngrams = 2
)
```

## Typical workflow

A common pattern is to iterate with `edstr_view()` until the regex captures the right matches, then use the finalised pattern in `edstr_extract()`:

1. Start broad: `edstr_view(data, pattern = "diabet", ngrams = 3)`
2. Review the `count` table to identify common expressions and false positives
3. Refine the pattern or add exclusions
4. Once satisfied, pass the pattern as a concept to `edstr_extract()`
