---
title: "edstr_extract(): text extraction"

vignette: >
  %\VignetteIndexEntry{edstr_extract(): text extraction}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
library(edstr)
```

`edstr_extract()` is the core of the pipeline.
It tokenises cleaned text, matches user-defined concepts via regex, filters false positives, and exports results as XLSX, CSV, and RDS files.

## Prerequisites

```{r}
#| error: true
edstr_config(
  edstr_dirname = "output/my_study",
  edstr_filename = "my_study",
  edstr_text = "note_text",
  edstr_overwrite = FALSE
)

df_import <- edstr_import()
df_clean <- edstr_clean(data = df_import, replace = c("\\s+" = " "))
```

## Defining concepts

Concepts are named regex patterns that define the clinical entities to search for.
A named character vector creates flat, independent concepts:

```{r}
#| label: extract-flat
#| error: true
result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture", femur = "f(e|é)mur|fesf"),
  group = "id_pat"
)
```

A nested named list creates sub-concepts grouped under a root:

```{r}
#| label: extract-nested
#| error: true
result <- edstr_extract(
  data = df_clean,
  concepts = list(
    fracture = c(
      fesf = "fesf|extremite superieure",
      col  = "col (du )?femur"
    )
  ),
  group = "id_pat"
)
```

## Tokenisation

The `token` argument controls the n-gram sizes used for tokenisation.
Unigrams (`token = 1`) are the default; adding bigrams captures multi-word expressions.

```{r}
#| label: extract-bigrams
#| error: true
result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  token = c(1, 2),
  group = "id_pat"
)
```

By default, `starts_with_only = TRUE` appends `\S*$` to patterns, matching any token that starts with the concept.
Set `starts_with_only = FALSE` for exact matching.

## Collapse and intersect

When multiple concepts are defined, two modes alter the matching logic:

- **`collapse = TRUE`**: OR-collapses all concept patterns into a single regex per root concept.
  Useful when sub-concepts are synonyms that should be treated as one.
- **`intersect = TRUE`**: keeps only documents that match **all** root-level concepts.
  Useful for narrowing results to co-occurrences.

Both require at least two concepts.

```{r}
#| label: extract-intersect
#| error: true
result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture", femur = "f(e|é)mur|fesf"),
  intersect = TRUE,
  group = "id_pat"
)
```

## Exclusions

False positives can be removed through two mechanisms:

### Manual exclusions

A regex pattern passed to `exclus_manual` removes any matched token containing that pattern.

```{r}
#| label: extract-exclus-manual
#| error: true
result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  exclus_manual = "fracture ouverte|ancienne fracture",
  group = "id_pat"
)
```

### Automatic exclusions

Heuristic-based filtering on long tokens.
`exclus_auto_token_min` sets the minimum n-gram size (default `10`) above which auto-exclusion applies.
`exclus_auto_escape` removes specific tokens from the match pool before auto-exclusion runs.

## Sampling

For development and testing, `sample` draws a random subset of rows before extraction.
Use `seed` for reproducibility.

```{r}
#| label: extract-sample
#| error: true
result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  sample = 500,
  seed = 42,
  group = "id_pat"
)
```

## Pseudonymisation

Two arguments handle de-identification before extraction:

- `ano_hash`: column names whose values are replaced by truncated hashes.
- `ano_hide`: column names whose values are masked with `"---"`.

```{r}
#| label: extract-ano
#| error: true
result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  ano_hash = "id_pat",
  ano_hide = "patient_name",
  group = "id_pat"
)
```

## Output structure

`edstr_extract()` returns a nested list and saves three files:

| File    | Contents |
|---------|----------|
| `.xlsx` | Excel workbook with one sheet per result type (extraction, token counts, exclusions, source matching, mismatch, parameters) |
| `.csv`  | Flat extraction table for downstream analysis |
| `.rds`  | Full list with all intermediate objects |

The returned list contains:

- `data`: data frames (`base`, `match`, `extract`, `csv`)
- `regex`: parsed patterns, replacement rules, source-level matches
- `match`: initial and final (post-exclusion) matches
- `count`: token-level and distinct match counts
- `exclus`: excluded matches and counts
- `mismatch`: discrepancies between token and source matching
- `summary`: summaries by token and concept, plus call parameters
- `sheets`: data frames and optional `gt` tables for each Excel sheet
