---
title: "edstr_extract(): text extraction"

vignette: >
  %\VignetteIndexEntry{edstr_extract(): text extraction}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup

library(edstr)
```

`edstr_extract()` is the core of the pipeline.
It tokenises cleaned text, matches user-defined concepts via regex, filters false positives, and exports results as XLSX, JSON, and RDS files.

## Prerequisites

```{r}
#| error: true

edstr_config(
  edstr_dirname = "output/my_study",
  edstr_filename = "my_study",
  edstr_text = "note_text",
  edstr_overwrite = FALSE
)

df_import <- edstr_import()
df_clean <- edstr_clean(data = df_import, replace = c("\\s+" = " "))
```

## Defining concepts

Concepts are named regex patterns that define the clinical entities to search for.
A named character vector creates flat, independent concepts:

```{r}
#| label: extract-flat
#| error: true

result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture", femur = "femur|fesf"),
  group = "id_pat"
)
```

A nested named list creates sub-concepts grouped under a root:

```{r}
#| label: extract-nested
#| error: true

result <- edstr_extract(
  data = df_clean,
  concepts = list(
    fracture = list(
      fesf = "fesf|extremite superieure",
      col = "col (du )?femur"
    )
  ),
  group = "id_pat"
)
```

## Tokenisation

The `token` argument controls the n-gram sizes used for tokenisation.
Unigrams (`token = 1`) are the default; adding bigrams captures multi-word expressions.

```{r}
#| label: extract-bigrams
#| error: true

result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  token = c(1, 2),
  group = "id_pat"
)
```

By default, `starts_with_only = TRUE` appends `\S*$` to patterns, matching any token that starts with the concept.
Set `starts_with_only = FALSE` for exact matching.

## Collapse and intersect

When multiple concepts are defined, two modes alter the matching logic:

- **`collapse = TRUE`**: OR-combines patterns into a single regex, one for the whole set when concepts are flat, one per root concept when they are nested.
  Useful when sub-concepts are synonyms that should be treated as one.
- **`intersect = TRUE`**: keeps only documents that match **all** root-level concepts.
  Useful for narrowing results to co-occurrences.

Both require at least two concepts.

```{r}
#| label: extract-intersect
#| error: true

result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture", femur = "femur|fesf"),
  intersect = TRUE,
  group = "id_pat"
)
```

## Exclusions

False positives can be removed through two mechanisms:

### Manual exclusions

A regex pattern passed to `exclus_manual` removes any matched token containing that pattern.

```{r}
#| label: extract-exclus-manual
#| error: true

result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  exclus_manual = "fracture ouverte|ancienne fracture",
  group = "id_pat"
)
```

### Automatic exclusions

Heuristic-based filtering on long tokens.
`exclus_auto_token_min` sets the n-gram size above which auto-exclusion applies, in the same unit as `token`.
Its default of `10` is above any realistic n-gram size, so out of the box the heuristic scans and then discards everything it found: lower the threshold below your smallest `token` value to actually enable it.
`exclus_auto_escape` removes specific tokens from the match pool before auto-exclusion runs.

## Sampling

For development and testing, `sample` draws a random subset of rows before extraction.
Use `seed` for reproducibility.

```{r}
#| label: extract-sample
#| error: true

result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  sample = 500,
  seed = 42,
  group = "id_pat"
)
```

## Pseudonymisation

Two arguments handle de-identification before extraction:

- `ano_hash`: column names whose values are replaced by truncated hashes.
- `ano_hide`: column names whose values are masked with `"---"`.

```{r}
#| label: extract-ano
#| error: true

result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  ano_hash = "id_pat",
  ano_hide = "patient_name",
  group = "id_pat"
)
```

## Unmatched documents

Documents that produce no token match are reported separately, split by reason:

- `no_concept`: the text was searched and matched no concept.
- `no_source`: the source is empty or missing, so there was nothing to search.
- `empty_text`: the source holds no text once its markup is stripped.
- `outside_p`: the document has text, but it falls outside the `<p>` blocks the formatter reads.

Only `no_concept` is a result: it is the set of true negatives, the one to use as a denominator.
The other three record documents the pipeline could not evaluate, and keeping them apart is what makes that denominator trustworthy.

The extraction summary always reports how many documents produced no matchable text, splitting the `no_source`, `empty_text` and `outside_p` counts.
Those three sets are always populated in the output.
The `no_concept` set can be large, so its rows are materialised only when `unmatched_data = TRUE`.
Its count, `unmatched$n_no_concept`, is always exact regardless: read it for the denominator, never `nrow(unmatched$no_concept)`, which is `0` both when the set is genuinely empty and when it is merely unmaterialised.

```{r}
#| label: extract-unmatched
#| error: true

result <- edstr_extract(
  data = df_clean,
  concepts = c(fracture = "fracture"),
  unmatched_data = TRUE,
  group = "id_pat"
)
```

## Output structure

`edstr_extract()` returns a nested list and saves three files:

  | File    | Contents                                                                                                                                                                                                                                                    |
  | ------- | ----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
  | `.xlsx` | Excel workbook with one sheet per result type: the extraction, the token stage (patterns, matches, counts, exclusions), concept counts, the source-matching stage (replacements, patterns, matches, counts), unmatched, mismatched, and the call parameters |
  | `.json` | Summary tables (by token, by concept) and the call parameters (JSON format)                                                                                                                                                                                 |
  | `.rds`  | Full list with all intermediate objects (R-native, used for caching)                                                                                                                                                                                        |

The returned list contains:

- `data`: data frames (`base`, `match`, `extract`, `note`)
- `regex`: parsed patterns, replacement rules, source-level matches
- `match`: initial and final (post-exclusion) matches
- `count`: token-level and distinct match counts
- `exclus`: excluded matches and counts
- `unmatched`: the true-negative count (`n_no_concept`, always exact) and documents with no token match, split by reason (`no_concept`, `no_source`, `empty_text`, `outside_p`)
- `mismatched`: discrepancies between token and source matching
- `summary`: summaries by token and concept, plus call parameters
- `sheets`: data frames and optional `gt` tables for each Excel sheet
