---
title: "edstr_import(): importing data"

vignette: >
  %\VignetteIndexEntry{edstr_import(): importing data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
library(edstr)
```

`edstr_import()` executes a SQL query against an Oracle database and saves the result as an RDS file.
It requires `edstr_config()` to be called first.

## Prerequisites

```{r}
edstr_config(
  edstr_dirname = "output/my_study",
  edstr_filename = "my_study",
  edstr_text = "note_text",
  edstr_overwrite = FALSE
)
```

## Running a query

The `query` argument accepts either a SQL string or a path to a `.sql` file.
Connection parameters (`user`, `password`, `tns`) configure access to the Oracle database.

```{r}
#| error: true
df_import <- edstr_import(
  query = "SELECT * FROM clinical_notes WHERE rownum <= 1000",
  user = "my_user"
)
```

Using a `.sql` file keeps queries out of R scripts and makes them easier to version:

```{r}
#| error: true
df_import <- edstr_import(
  query = "sql/clinical_notes.sql",
  user = "my_user"
)
```

## Limiting rows

The `head` argument appends `FETCH FIRST ... ROWS ONLY` to the query.
This is useful for development and testing without modifying the SQL itself.

```{r}
#| error: true
df_import <- edstr_import(
  query = "sql/clinical_notes.sql",
  head = 500,
  user = "my_user"
)
```

## Loading cached data

When the output file already exists, `edstr_import()` skips the database query entirely and loads from cache.
The behaviour depends on the `edstr_overwrite` option set in `edstr_config()`.

```{r}
#| error: true
df_import <- edstr_import()
```

::: {.callout-note}
## Column names

By default, `lower = TRUE` converts all column names to lowercase after import.
Set `lower = FALSE` to preserve the original casing from the database.
:::
