---
title: "Matching: tokens, accents, and the source text"

vignette: >
  %\VignetteIndexEntry{Matching: tokens, accents, and the source text}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup

library(edstr)
```

`edstr_extract()` matches twice, against two different representations of the same document.
Knowing which is which explains why concept patterns are written without accents, and how the original spelling is restored in the results.

## Two representations, two matching stages

  | Stage        | What is matched         | Against what                                               |
  | ------------ | ----------------------- | ---------------------------------------------------------- |
  | Match token  | your concept patterns   | the tokenised text: HTML stripped, transliterated to ASCII |
  | Match source | the tokens that matched | the raw source text, accents and tags intact               |

`edstr_extract()` first formats the text: it keeps the content of `<p>` blocks, drops the remaining tags, and transliterates the result with `Latin-ASCII`.
That formatted text is what gets tokenised, so **every token in the corpus is ASCII**.

```{r}
#| eval: true

stringi::stri_trans_general("œdème méningé", "Latin-ASCII")
```

This is why concepts are written without accents.
A pattern containing `"diabète"` cannot match a token vocabulary in which the word is stored as `diabete`.

```{r}
result <- edstr_extract(
  data = df_clean,
  concepts = c(diag = "diabete|oedeme", sympt = "cephalee")
)
```

## Finding the token again in the source

The second stage has to locate each matched token in the raw text, so that the extracted result shows what the clinician actually wrote rather than an ASCII approximation.

It does this by applying the **same** `Latin-ASCII` transliteration to the source before searching.
Token and folded source are then the same string by construction, so nothing has to bridge a spelling gap between them.

```{r}
#| eval: true

source_text <- "patiente avec œdème méningé et cœur normal"
folded <- stringi::stri_trans_general(source_text, "Latin-ASCII")
folded
```

The match is found on the folded text, but the value returned is sliced from the original.
Accents are never lost, because they are never removed from what the function gives back.

## Carrying positions back to the original

`Latin-ASCII` does not preserve string length: the oe-ligature is one character that transliterates to two.
Positions found on the folded text therefore need mapping back, which `edstr_extract()` does in two steps: from the byte offsets the matching engine reports to character indices in the folded text, then from those to character indices in the original.

```{r}
#| eval: true

map_back <- \(pos, width) findInterval(pos - 1L, cumsum(width)) + 1L

chars <- stringi::stri_sub(source_text, seq_len(nchar(source_text)), length = 1)
width <- nchar(stringi::stri_trans_general(chars, "Latin-ASCII"), "chars")

found <- stringr::str_locate_all(folded, "oedeme|coeur")[[1]]

stringi::stri_sub(
  source_text,
  map_back(found[, "start"], width),
  map_back(found[, "end"], width)
)
```

Both terms come back in their original spelling, ligature included, even though the pattern that found them was plain ASCII.

The mapping assumes the transliteration is context free, meaning that transliterating a string character by character gives the same result as transliterating it whole.
`edstr_extract()` checks this per document and falls back to a slower accent-aware search when it does not hold.

## Why not simply strip the accents

A tempting alternative is a fold that only removes diacritics, by decomposing each character, dropping the combining marks, and recomposing.
It has the appeal of preserving string length on precomposed text, which makes position mapping trivial there.
Text that arrives already decomposed shortens instead, since the marks it drops were separate characters to begin with, so equal length is never on its own a proof that positions did not move.

It also silently loses the ligatures, because Unicode gives them no decomposition at all.

```{r}
#| eval: true

strip_marks <- \(x) {
  stringi::stri_trans_general(x, "NFD; [:Nonspacing Mark:] Remove; NFC")
}

c(accent = strip_marks("méningé"), ligature = strip_marks("cœur"))
```

Unlike the typographic `ﬁ` ligature, `œ` and `æ` are letters in their own right in the Unicode character database, carrying neither a canonical nor a compatibility decomposition.
No normalization form separates them.

```{r}
#| eval: true

c(fi = stringi::stri_trans_nfkd("ﬁ"), oe = stringi::stri_trans_nfkd("œ"))
```

Clinical French uses those letters in common terms, `cœur`, `œdème`, `œsophage`, `cœlioscopie`, so a length-preserving fold would leave them unmatchable.
Paying for an offset map buys their coverage.

## The `regex_replace` argument

Independently of the fold, the source-matching pattern expands each vowel into a character class covering its accented variants.
This is what the accent-aware fallback search relies on, and it is the extension point you can add to.

  | Written | Matches                                              |
  | ------- | ---------------------------------------------------- |
  | `oe`    | `[oô][eéèêë]` or the `œ` ligature                    |
  | `ae`    | `[aàâ][eéèêë]` or the `æ` ligature                   |
  | `a`     | `[aàâ]`                                              |
  | `e`     | `[eéèêë]`                                            |
  | `i`     | `[iîï]`                                              |
  | `o`     | `[oô]`                                               |
  | `u`     | `[uùûü]`                                             |
  | space   | whitespace, hyphens, apostrophes, and `<br/>` breaks |

A ligature needs the two-letter rules: the fold expands it, so it reaches a token as the pair it stands for and a single-letter class could never match it back.

Pass `regex_replace` to `edstr_extract()` to add rules, for instance to let a term match a spelling variant specific to your corpus.
Every element must be named, the name being the pattern its value replaces, and the value reaches the source pattern as the regex fragment you write.
The built-in rules run first and their output is inert, so a rule whose pattern contains a character the table above already covers never fires: `"y"` is a usable rule, `"oe"` is not.
A rule whose pattern reaches any character (`.`, `\W`, a negated class) reaches that same machinery and is rejected rather than applied.
A backreference in a replacement is not supported either.

## What this means in practice

- Write concepts against ASCII: `"oedeme"`, not `"œdème"`.
  Both spellings in the source are found, and each is returned as written.
- Extracted text keeps its original spelling, accents, ligatures and case included.
- Extend `regex_replace` when a term varies in ways the default classes do not cover.
- The `mismatched` output lists tokens that matched the tokenised text but could not be confirmed in the source.
  A recurring entry there usually points at a formatting rule, not at a concept pattern.
